Owlifier: Creating OWL-DL Ontologies from Simple Spreadsheet-Based Knowledge Descriptions\tnoteref{t1}

### 2009-08-29 docs/pubs/ISEI6/response.txt

We thank the reviewers for their excellent comments and suggestions for improvements to the paper.  We revised the paper to address each of the suggestions. Below we describe the changes we made that were requested by the reviewers as well as answers to the questions of the reviewers.

@@

---------------------------------------------------------------------- Reviewer #1: ---------------------------------------------------------------------- 1). "Not mentioned is the list of OWL features not supported."

We added a sentence to the implementation section clarifying this, and also a brief description of what is not supported to the introduction (along with some motivation for why we consider only a subset of constructs).

---------------------------------------------------------------------- 2). "Overriding the OWL open world assumption should be fine, so long as that assumption is made clear to the users in the documentation."

We primarily provide a different approach (e.g., than Protege) for specifying disjoint classes, as opposed to a general override of the OWA. However, we agree that describing this feature will be particularly important to users of owlifier that are familiar with OWL-DL. We also describe this in Section 3.1.

@@

---------------------------------------------------------------------- 3). "There is no mention of the competing CMAP-COE tool that makes it easy to create ontologies visually, with help from drop-down lists.  It would be helpful to include a line that compares your approach to CMAP (which also supports only a subset of OWL)."

We added additional text in the introduction that further describes our motivation for the spreadsheet-based approach and some of its advantages for ecologists/scientists over CMAP-COE. This is also now briefly mentioned in the conclusion (which lists additional related work).

@@

---------------------------------------------------------------------- 4). "Something that might be pointed out - the tool could be used to help enforce some community standards (e.g., a common use of a "part-of" relation)."

We added two sentences to the conclusion that highlights this feature of owlifier.

@@

---------------------------------------------------------------------- 5). "In the references: Usually, the word "pages" is abbreviated to pp. Ref 18: should be to a journal article, not a web site - perhaps Raskin and Pan (2005) in Computers & Geosciences. Ref 19: to EKAW?  What is that?  Compare Refs 4 and 7 - different wordings are used to describe the same conference."

We fixed each of these.

---------------------------------------------------------------------- Reviewer #2: ---------------------------------------------------------------------- 6). "In the early stages of the paper there are some assertions without reference about what is and is not difficult. See p.2 end of 2nd, 3rd and 4th paragraphs.  Can a little more detail be added to support the statements? If not, then noting that these are suppositions would be useful for the reader."

To address this issue we added additional text in the introduction on why there is a need for new tools familiar to ecological scientists for ontology development, comparing our approach to CMAP-COE and Protege as well as efforts such as OBO.  Also, much of the motivation for the work is based on our work with ecologists and information managers, which we refer to in the paper (in the implementation section).

@@

---------------------------------------------------------------------- 7). "Was there any quantifiable basis for choosing the predicates?  In particular, the statement" "While not as expressive as OWL-DL, our approach can produce ontology structures that are essential for improved data discovery and integration [14]."  indicates some level of decision in the reduced description but without any details, especially in terms for DL-expressivity, it is SHIQ, SHOIN, or what?"

As far as we are aware, the arrangement of DL constructs used in owlifier does not directly map to any previously named DL sublanguage, however, it does provide only OWL-DL based constructs, and thus can be considered best as a sublanguage of SHOIN(D). Note that although many of the OWL-DL constructs can be used, they can only be used in certain contexts. Also, some of the more foundational constructs of DLs are not considered, e.g., value restrictions and concept disjunction. While additional blocks can easily be added to owlifier spreadsheet conventions, the proposed set of blocks were chosen to provide a simplified syntax (for a novice DL user) while leading to a "natural" or "intuitive" semantics of blocks for these users.

@@

---------------------------------------------------------------------- 8). "There are no mention of instances, that I could find, at all of instances?  Are these part of the use case or not? If not, this really needs to be stated.  An ontology without instances has limited application."

Our main goal is to provide a mechanism for describing ecological concepts, and then to allow additional tools to use these concepts for annotating scientific data (which represents the instances/values). We briefly mention this now in the introduction, and also elaborate on the scope of owlifier later in the paper. Also see our response to (14) below.

---------------------------------------------------------------------- 9). "Fig.1 suggests a round-trip approach to knowledge engineering, but in fact it is not. How does the present formulation represent to scientist what gets dropped in the back conversion from OWL to the spreadsheet format?"

We reworded the implementation section to make it clearer that the conversion is not information preserving.  In general, the owlifier tool will only notify the user of this situation, but at this time no additional support is provided. We also tried to better clarify and describe the extensions to owlifier needed to better support the round trip conversion in the conclusion (as part of our future work).

---------------------------------------------------------------------- 10). "Most of all, there is almost no (or none at all) description of the implementation or any evaluation, i.e. the how, why, etc.  What has this application been applied to?"

The areas the owlifier application has been applied to is described in the implementation section, as well as in the introduction (briefly). In particular, we are applying owlifier to the Traitnet project (ecologists studying/using biological trait data) and to an LTER project that is attempting to impose structure over a large list of keywords extracted from existing metadata documents. We have not performed a formal evaluation of the owlifier application, which we state in the paper as future work, however, we have received positive (informal) feedback on the tool and we briefly describe this at the end of the implementation section.

@@

---------------------------------------------------------------------- 11). "It seems that it would be easy to generate complex blocks. How are moderate size ontologies developed?"

Our use of owlifier has included both small and moderate sized ontologies.  Ecologists often create spreadsheets with large amounts of data, and similarly owlifier can be used in this way. In addition, sparse blocks significantly simplify the size of large hierarchies, since not all information needs to be repeated per row. In the LTER project mentioned above, there are thousands of terms (organized into a flat list) that are easily imported into Excel, and then using the Excel editing features, have been fairly easily organized into hiearchies (i.e., entity blocks and property blocks).  In general, using a combination of tools (e.g., that could show a portion of the spreadsheet visually) would also probably be extremely useful, however, we have not developed this type of technology for owlifier.

@@

---------------------------------------------------------------------- 12). "What is the rationale for the proposed extensions later in the paper, i.e. what are the use cases or evaluation-driven itersations?"

The point of these extensions are primarily for round-trip support of owlifier to OWL-DL and back. We added and revised the last paragraph to better articulate the goals and motivation of this work.

@@

---------------------------------------------------------------------- 13). "Possibly the authors could signal at the end of section 1 something like "Readers unfamiliar with logic notation may safely ignore the small amount they may encounter in what follows.""

We added a sentence stating this.

---------------------------------------------------------------------- 14). "Although not enough to recommend required additions, I am left a little puzzled as to why the authors seem to be silent on two issues: OWL data type properties and owlifier instance entities. Both of these would seem easy to represent in their owlifier syntax. I would also imagine that both could arise in representing and reasoning about both dataset metadata and record-level metadata on biodiversity databases. So I am left wondering whether I am wrong about the ease of representing them in spreadsheets and attaching DL axioms, or whether instead, the authors believe that these two OWL constructs are better treated in as data for reasoners or elsewhere."

Our use cases are largely driven by separating the specification of conceptual information (i.e., t-box assertions) from instance-level information.  In particular, in our applications instance level information is provided through "semantic annotations" that map data sets to ontology instances.  In this way, too, multiple data sets can be mapped to the same ontologies. Our decision also is inspired in part by the work of Rector et al.  that state in the OWL Pizza's paper: "this paper only concerns issues in defining OWL classes, since this is the strength of OWL-DL, and most existing classifiers deal with individuals completely or not at all."  However, we may extend owlifier in the future to include datatype properties, which would be a natural and simple extension.

@@

---------------------------------------------------------------------- 15). "On page 4 (Subsection "Overlap blocks", the second sentence has a small English ambiguity: "[...]a given set of classes may have overlapping instances." could reasonably be understood to mean that some instances overlap one another, whereas I believe the clause means "several classes may have common instances." [Grammatically, the overlapping instances do not belong to the collective "set of classes" but rather to each of the members of that collective]"

We fixed this, which now reads: "Overlap blocks explicitly relax this assumption by stating that a given set of classes may share common instances."

### 2009-08-29 docs/pubs/ISEI6/main.tex
almost done...

\documentclass[5p,authoryear]{elsarticle} 
% other options: preprint, review, 1p, 3p, 5p, 
% bib options: authoryear, number, longtitle,
% other options: times
\usepackage{graphicx}

@@

\address[smb]{UC Davis Genome Center \& Dept. of Computer Science,
  Gonzaga University}

@@

Efforts to engage scientists in the development of ontologies
typically leverage the W3C Web Ontology Language (OWL)
\citep{smith04:_owl_web_ontol_languag_guide} as a standard XML syntax
for representing and sharing ontologies. A key advantage of OWL is
that it is supported by a wide range of generic tools, including
editors
\citep{knublauch04:_editin_descr_logic_ontol_with,kalyanpur05:_swoop},
reasoning systems
\citep{sirin07:_pellet,tsarkov06:_fact_descr_logic_reason}, query
languages
\citep{prudhommeaux08:_sparq_query_languag_for_rdf,motik05:_query_answer_for_owl_dl_with_rules},
and storage technologies
\citep{carroll04:_jena,broekstra02:_sesam}. Most of these tools,
however, are primarily targeted at experts in knowledge engineering
and software development familiar with the underlying description
logic semantics of OWL-DL \citep{grau08:_owl}. This is especially true
with ontology editors (such as Protege, SWOOP, etc.), which allow for
very detailed ontology specifications, but at the same time require
considerable understanding of the underlying ontology formalisms and
syntax.  A well-funded ontology development effort, the Gene Ontology
(GO), has been underway for over a decade within the molecular biology
community, and has led to some major successes enriching genomic and
biomedical data resources with annotations to community-based
ontologies. A major lesson learned from these efforts was the
importance of keeping the ontology focused, and indeed GO has a highly
delimited scope \citep{bada04:_short_study_succes_of_gene_ontol}
Ecology, on the other hand, encompasses an extremely broad range of
scales and disciplines, such that a coordinated ontology development
effort would need to involve a large number of experts to cover the
diverse types of information relevant to ecological investigations.
At this point, we believe that ecological researchers need to begin
constructing ontologies for their own specialized subdisciplines.
Most ontology editing tools, however, typically require substantial
expertise in formal knowledge representation, and present a major
barrier for more wide-scale creation and adoption of ontologies in
ecology.

This paper presents a novel approach for ontology creation that aims
at being more intuitive for ecologists and that can be used to rapidly
construct large ontologies of ecological terms (classes) and the
subsequent annotation of scientific data (instances).  Our approach is
to allow scientists to use common and familiar spreadsheet-based tools
(e.g., Microsoft Excel, Apple Numbers, Open Office Spreadsheet) to
describe, in an intuitive way, different aspects of an ontology, and
then to take these descriptions and convert them into full-fledged OWL
ontologies using a software application called \owlifier. The use of a
\emph{familiar} editing environment can improve the speed and
understanding of ontology construction and can help users avoid the
often time-consuming task of locating, downloading, installing, and
learning fundamentally new software applications and interfaces (e.g.,
CMAP-COE \citep{hayes05:_collab_knowl_captur_in_ontol}, Protege
\citep{knublauch04:_editin_descr_logic_ontol_with}).  Many spreadsheet
applications also contain easy-to-use features that support scientists
in the process of editing and managing \owlifier\ ontology documents
(e.g., via cut/copy/paste, search/replace, sort, split-window,
track-changes, freeze panes, and so on).

An \owlifier\ spreadsheet consists of a set of \emph{blocks} that have
a predefined template structure for users to fill in. Each non-empty
row in an \owlifier\ table constitutes a block. Each block defines
different aspects of an ontology including ontology classes,
subclasses, synonyms, and properties.  We also provide blocks for
plain-text descriptions of classes and properties, and for referencing
one or more existing ontologies (e.g., to extend an existing ontology
or to define ontology articulations). Blocks can be sparse (inheriting
from previous blocks), which can further simplify the creation of
large ontologies. We omit several OWL features associated with
properties (e.g., role inclusion axioms, domain and range property
restrictions, and data-type properties among others) as these can be
confusing to construct for non-experts and are more suitable for
experienced knowledge modelers.

While not as expressive as OWL-DL, our approach can produce ontology
structures that are essential for improved data discovery and
integration \citep{madin07:_ontol_for_descr_and_synth}. Just as
important, because spreadsheet tools are frequently used by ecologists
to store and analyze data, \owlifier\ can provide ecologists with a
familiar and accessible user interface for ontology creation. This
approach also leverages the ability of spreadsheet tools to organize
and manipulate tabular data, e.g., allowing users to rapidly construct
class hierarchies from long lists of keywords.  In this way, an
ecologist can easily construct (or import) a set of terms, and then
incrementally organize these into class hierarchies, properties, and
constraints. In initial experiments with ecologists and evolutionary
biologists studying trait data, we found that \owlifier\ enabled them
to quickly and easily comprehend and construct relatively complex and
meaningful ontologies.

The rest of this paper is organized as follows. In
\secref{sec:owlifier} we describe the basic syntax and semantics of
\owlifier. We define blocks that support a large subset of OWL-DL and
that also generally follow the ontology creation guidelines defined in
\citep{rector04:_owl_pizzas}. We also simplify certain aspects of
ontology creation using OWL-DL, e.g., by assuming classes are disjoint
by default (unless specified otherwise)
\citep{rector04:_owl_pizzas}. In \secref{sec:features} we describe
additional characteristics of \owlifier\ and discuss issues with
respect to classification and reasoning. In
\secref{sec:implementation} we briefly describe the \owlifier\
implementation, and conclude in \secref{sec:conclusion} with related
and future work. In general, the goal of \owlifier\ is not to support
all constructs in OWL-DL, but instead to provide a higher-level
ontology syntax (via spreadsheet blocks) that is easy for ecologists
to use and understand while also providing the necessary constructs
for developing typical ecological ontologies. By compiling \owlifier\
to OWL-DL, we also allow for experts to refine and extend the ontology
using more advanced ontology editing tools if necessary
(cf.~\figref{fig:owlifier}).

@@

As described above, an \owlifier\ table defines an OWL-DL
\citep{smith04:_owl_web_ontol_languag_guide} ontology through a set of
\emph{blocks} representing one or more ontology definitions.  Each
non-empty row in an \owlifier\ table corresponds to a block. The type
of the block is given in the first column of the row. 
% This section describes in detail each type of block supported by
% \owlifier.
We assume that if any properties or classes used in a block are not
imported from another ontology, then they are to be added to the
ontology being specified by the \owlifier\ table (i.e., the
``current'' ontology). In general, we name blocks according to the
terms used in
\citep{bowers08:_concep_model_framew_for_expres,madin07:_ontol_for_descr_and_synth}
as opposed to the names used for corresponding constructs in
OWL-DL. This choice of block names helps to simplify terminology
(e.g., we use ``relation'' below instead of ``object property''),
allows \owlifier\ to easily generate ontologies that extend the
observational model of
\citep{bowers08:_concep_model_framew_for_expres,madin07:_ontol_for_descr_and_synth},
and avoids confusion with established terms commonly used within
ecology (e.g., ``class'').

@@

\myblock{Overlap Blocks.} Except in certain situations (described
further in \secref{sec:features}), classes are generally
assumed to be disjoint in \owlifier.  Overlap blocks explicitly relax
this assumption by stating that a given set of classes may have
overlapping instances. Rows containing overlap blocks take the form
\begin{itemize}
\item[] 
  \begin{tabular}{|l|l|l|l|l|}\hline
    \textsf{overlap} & $c_1$ & $c_2$ & \dots & $c_n$ \\ \hline 
  \end{tabular} \hfill ($n \ge 2$)
\end{itemize}
where each class $c_i$ is allowed to share instances with each class
$c_j$, for $1 \le i,j \le n$. That is, a given $c_i$ and $c_j$ in an
overlap block are not defined to be disjoint classes in the current
ontology. As an example, consider the following entity blocks that
define the classes Estuary, Lagoon, and Marsh as subclasses of
Ecological Habitats.
\begin{tabbing}
  ~~\textsf{entity} \textsf{EcologicalHabitat} \textsf{Estuary} \\ 
  ~~\textsf{entity} \textsf{EcologicalHabitat} \textsf{Lagoon} \\ 
  ~~\textsf{entity} \textsf{EcologicalHabitat} \textsf{Marsh} 
\end{tabbing}
Given only these blocks, \owlifier\ treats Estuary, Lagoon, and Marsh
as disjoint classes. To relax this assumption and allow, e.g., types
of Lagoons to also be types of Estuaries, we explicitly add the
following overlap block
\begin{tabbing}
  ~~\textsf{overlap} \textsf{Estuary} \textsf{Lagoon}
\end{tabbing}
In general, overlap blocks are rarely used but provide a mechanism to
override the default behavior of \owlifier\ in asserting disjoint
classes.

@@

\myblock{Sufficient Blocks.} Sufficient blocks are similar to synonym
blocks in that they state equivalences between classes. A sufficient
block serves as a modifier to entity blocks, relationship blocks
(including transitive blocks), and cardinality blocks, modifying the
associated description logic axioms by using equivalence ($\equiv$) in
place of subclass ($\sqsubseteq$). For example the block
\begin{tabbing}
  ~~\textsf{sufficient} \textsf{relationship} \textsf{hasPart} \textsf{Mammal} \textsf{Hair} 
\end{tabbing}
induces the description logic axiom $\textsf{Mammal} \equiv \exists
\textsf{hasPart}.\textsf{Hair}$ stating that \emph{any} individual
that has Hair as a part is a Mammal. Additionally, we allow
relationship blocks within a sufficient block to be extended with
\textsf{not} to state that the absence of the property is a defining
characteristic of the class. For example the following blocks
\begin{tabbing}
  ~~\textsf{sufficient} \textsf{entity} \textsf{Mammal} \textsf{Eutheria} \\
  ~~\textsf{sufficient} \textsf{relationship not} \textsf{hasPart} \textsf{Eutheria} \textsf{EpipubicBone} 
\end{tabbing}
induces the axiom $\textsf{Eutheria} \equiv \textsf{Mammal} \sqcap
\neg \exists \textsf{hasPart} . \textsf{EpipubicBone}$ stating that
any Mammal that does not have an Epipubic Bone as a part is a
Eutheria. Note also that multiple sufficient blocks for a particular
class result in a single axiom in which each constraint (i.e., class
constructor) is combined via intersection ($\sqcap$).

@@

\figref{fig:owlifier} shows the general architecture of the \owlifier\
application. A scientist or researcher first creates a spreadsheet
containing a set of ontology definitions using \owlifier\ conventions.
The spreadsheet is then exported as a plain-text file containing the
\owlifier\ table, e.g., using a CSV or tab-delimited format. The
\owlifier\ table is then sent to the \owlifier\ application, which
performs a number of steps that include: (i) parsing the file; (ii)
generating the corresponding OWL-DL representation (e.g., in the
current implementation, the Jena API is used \citep{carroll04:_jena});
(iii) validating the ontology and ensuring it is consistent via an
OWL-DL reasoner (e.g., the current implementation uses Pellet
\citep{sirin07:_pellet}); and (iv) outputting the corresponding OWL-DL
file. The resulting OWL file can then be refined and extended via
existing OWL tools (e.g., Protege or SWOOP). It is possible that the
refined ontology is converted back to a corresponding \owlifier\
representation (shown using dashed lines in
\figref{fig:owlifier}). While straightforward to perform, this back
conversion is not information preserving since not all OWL-DL
statements can be represented through \owlifier\ blocks.

The current implementation of the \owlifier\ application is written in
Java and supports a subset of the blocks defined in
\secref{sec:owlifier}. In particular, we are currently extending
\owlifier\ to support cardinality and sufficient blocks as well as to
perform the (potentially lossy) back conversion from OWL-DL files to
corresponding \owlifier\ blocks. The current implementation of
\owlifier\ is being used within a project focused on integrating
vegetation trait data.  As noted above, not all the blocks described
here are yet supported, however, these priority needs are clearly
emerging from our interactions with the vegetation scientists.
Nevertheless, these scientists were able to rapidly prototype and
revise their trait ontologies using \owlifier, with little to no
instruction in formal logic or knowledge modeling.  In addition,
\owlifier\ is being used to construct term hierarchies from large sets
of keywords (harvested from existing metadata documents) as well as to
define articulations between existing ontologies and the observation
ontology defined in \citep{madin07:_ontol_for_descr_and_synth}.

@@

This paper presents a new approach for developing ontologies to
address barriers in ontology development and adoption by
ecologists. Our approach allows scientists to use familiar spreadsheet
software (e.g., frequently used by ecologists for storing and
analyzing data) to create ontologies by filling in a set of templates,
or blocks, that generate OWL-DL class hierarchies, properties, and
constraints via the \owlifier\ application.  This approach can provide
a more intuitive and accessible ontology editing environment for
ecologists, especially compared to existing OWL-based tools that
require considerable expertise in the underlying logic formalisms.

>>>>

\documentclass[5p,authoryear]{elsarticle} 
% other options: preprint, review, 1p, 3p, 5p, 
% bib options: authoryear, number, longtitle,
% other options: times
\usepackage{graphicx}
\usepackage{url}
\usepackage[colorlinks=true,hyperindex=false,bookmarks=false]{hyperref}

@@

\address[smb]{Dept. of Computer Science,
 Gonzaga University \& UC Davis Genome Center}

%\address[smb]{UC Davis Genome Center}

@@

Efforts to engage scientists in the development of ontologies typically leverage the W3C Web Ontology Language (OWL) \citep{smith04:_owl_web_ontol_languag_guide} as a standard XML syntax for representing and sharing ontologies. A key advantage of OWL is that it is supported by a wide range of generic tools, including editors \citep{knublauch04:_editin_descr_logic_ontol_with,kalyanpur05:_swoop}, reasoning systems \citep{sirin07:_pellet,tsarkov06:_fact_descr_logic_reason}, query languages \citep{prudhommeaux08:_sparq_query_languag_for_rdf,motik05:_query_answer_for_owl_dl_with_rules}, and storage technologies \citep{carroll04:_jena,broekstra02:_sesam}. However, most of these tools are primarily targeted at experts in knowledge engineering and software development familiar with the underlying description logic semantics of OWL-DL \citep{grau08:_owl}. This is especially true with ontology editors (such as Protege, SWOOP, etc.), which allow for very detailed ontology specifications, but at the same time require considerable understanding of the underlying ontology formalisms and syntax.  A well-funded ontology development effort, the Gene Ontology (GO), has been underway for over a decade within the molecular biology community, and has had success in enriching genomic and biomedical data resources with annotations to community-based ontologies. A major lesson learned from these efforts was the importance of keeping the ontology focused, and indeed GO has a highly delimited scope \citep{bada04:_short_study_succes_of_gene_ontol}. Ecology, on the other hand, encompasses an extremely broad range of scales and disciplines, such that a coordinated ontology development effort would need to involve a large number of experts to cover the diverse types of information relevant to ecological investigations.  At this point, we believe that ecological researchers need to begin constructing ontologies for their own specialized sub-disciplines, and due to the substantial expertise in formal knowledge representation required in most ontology editing tools, new tools are needed to enable wide-scale creation and adoption of ontologies in ecology.

This paper presents a novel approach for ontology creation that aims at being more intuitive for ecologists and that can be used to rapidly construct large ontologies of ecological terms (classes and properties), which can then be used in a subsequent step for annotation of scientific data (instances) \citep{bowers08:_concep_model_framew_for_expres,berkley09:_improv_data_discov_metad_repos_seman_searc}.  Our approach is to allow scientists to use common and familiar spreadsheet-based tools (e.g., Microsoft Excel, Apple Numbers, Open Office Spreadsheet) to describe, in an intuitive way, different aspects of an ontology, and then to take these descriptions and convert them into full-fledged OWL ontologies using a software application called \owlifier.

An \owlifier\ spreadsheet consists of a set of \emph{blocks} that have a predefined template structure for users to fill in. Each non-empty row in an \owlifier\ table constitutes a block. Each block defines different aspects of an ontology including classes, subclasses, synonyms, and properties.  We also provide blocks for plain-text descriptions of classes and properties, and for referencing one or more existing ontologies (e.g., to extend an existing ontology or to define ontology articulations). Blocks can be sparse (inheriting from previous blocks), which can further simplify the creation of large ontologies. We omit several OWL features associated with properties (e.g., role inclusion axioms, domain and range property restrictions, and data-type properties among others) as these can be confusing to construct for non-experts and are more suitable for experienced knowledge modelers.

While not as expressive as OWL-DL, our approach can produce ontology structures that are essential for improved data discovery and integration \citep{madin07:_ontol_for_descr_and_synth}. Just as important, because spreadsheet tools are already frequently used by ecologists to store and analyze data, \owlifier\ can provide ecologists with a familiar and accessible user interface for ontology creation. This approach also leverages the easy-to-use interfaces provided by many spreadsheet tools for organizing and manipulating tabular data, e.g., via cut/copy/paste, search/replace, sort, split-window, track-changes, freeze panes, and so on.  In this way, an ecologist can easily construct (or load) a set of terms, and then incrementally organize these into class hierarchies, properties, and constraints.  The use of an editing environment that is familiar to scientists can significantly help improve the speed and understanding of ontology construction and avoid the often time-consuming task of locating, downloading, installing, and learning fundamentally new software applications and interfaces (e.g., CMAP-COE \citep{hayes05:_collab_knowl_captur_in_ontol}, Protege \citep{knublauch04:_editin_descr_logic_ontol_with}). In initial experiments with ecologists and evolutionary biologists studying trait data, we found that \owlifier\ enabled them to quickly and easily comprehend and construct relatively complex and meaningful ontologies.

The rest of this paper is organized as follows. In \secref{sec:owlifier} we describe the basic syntax and semantics of \owlifier. The semantics of \owlifier\ blocks is given by mappings from \owlifier\ expressions to description-logic statements. Readers unfamiliar with description-logic notation may safely ignore these mappings, focusing instead on the descriptions and examples of blocks given in \secref{sec:owlifier}.  We define blocks that support a large subset of OWL-DL and that also generally follow the ontology creation guidelines defined in \citep{rector04:_owl_pizzas}. We also simplify certain aspects of ontology creation using OWL, e.g., by assuming classes are disjoint by default (unless specified otherwise) \citep{rector04:_owl_pizzas}. In \secref{sec:features} we describe additional characteristics of \owlifier\ and discuss issues with respect to classification and reasoning. In \secref{sec:implementation} we briefly describe the \owlifier\ implementation, and conclude in \secref{sec:conclusion} with related and future work. In general, the goal of \owlifier\ is not to support all constructs in OWL-DL, but instead to provide a higher-level ontology syntax (via spreadsheet blocks) that is easy for ecologists to use and understand while also providing the necessary constructs for developing typical ecological ontologies. By compiling \owlifier\ to OWL-DL, we also allow for experts to refine and extend the ontology using more advanced ontology editing tools if necessary (cf.~\figref{fig:owlifier}).

@@

As described above, an \owlifier\ table defines an OWL-DL \citep{smith04:_owl_web_ontol_languag_guide} ontology through a set of \emph{blocks} representing one or more ontology definitions.  Each non-empty row in an \owlifier\ table corresponds to a block. The type of the block is given in the first column of the row.
% This section describes in detail each type of block supported by
% \owlifier.
We assume that if any properties or classes used in a block are not imported from another ontology, then they are to be added to the ontology being specified by the \owlifier\ table (i.e., the ``current'' ontology). In general, we name blocks according to the terms used in \citep{bowers08:_concep_model_framew_for_expres,madin07:_ontol_for_descr_and_synth} as opposed to the names used for corresponding constructs in OWL-DL. This choice of block names helps to simplify terminology (e.g., we use ``relationship'' below instead of ``object property''), allows \owlifier\ to easily generate ontologies that extend the observational model of \citep{bowers08:_concep_model_framew_for_expres,madin07:_ontol_for_descr_and_synth}, and avoids confusion with established terms commonly used within ecology (e.g., ``class'').

@@

\myblock{Overlap Blocks.} Except in certain situations (described
further in \secref{sec:features}), classes are generally
assumed to be disjoint in \owlifier.  Overlap blocks explicitly relax this assumption by
stating that a given set of classes may share common instances. Rows containing overlap blocks take the form
\begin{itemize}
\item[]
  \begin{tabular}{|l|l|l|l|l|}\hline
    \textsf{overlap} & $c_1$ & $c_2$ & \dots & $c_n$ \\ \hline 
  \end{tabular} \hfill ($n \ge 2$)
\end{itemize}
where each class $c_i$ is allowed to share instances with each class
$c_j$, for $1 \le i,j \le n$. That is, a given $c_i$ and $c_j$ in an
overlap block are not defined to be disjoint classes in the current
ontology. As an example, consider the following entity blocks that
define the classes Estuary, Lagoon, and Marsh as subclasses of
Ecological Habitats.
\begin{tabbing}
  ~~\textsf{entity} \textsf{EcologicalHabitat} \textsf{Estuary} \\ 
  ~~\textsf{entity} \textsf{EcologicalHabitat} \textsf{Lagoon} \\ 
  ~~\textsf{entity} \textsf{EcologicalHabitat} \textsf{Marsh} 
\end{tabbing}
Given only these blocks, \owlifier\ treats Estuary, Lagoon, and Marsh
as disjoint classes. To relax this assumption and allow, e.g., types
of Lagoons to also be types of Estuaries, we explicitly add the
following overlap block
\begin{tabbing}
  ~~\textsf{overlap} \textsf{Estuary} \textsf{Lagoon}
\end{tabbing}
In general, overlap blocks are rarely used but provide a mechanism to
override the default behavior of \owlifier\ in asserting disjoint
classes.

@@

\myblock{Sufficient Blocks.} Sufficient blocks are similar to synonym blocks in that they state equivalences between classes. We consider two types of sufficient blocks. A sufficient entity block takes the form
\begin{itemize}
\item[]
  \begin{tabular}{|l|l|l|l|l|l|}\hline \textsf{sufficient} & \textsf{entity} & $c_1$ & $c_2$ & \dots & $c_n$
\\ \hline
\end{tabular} \hfill ($n \ge 2$)
\end{itemize}
where each $c_i$ is a class. A sufficient entity block induces the description logic axiom $c_1 \equiv c_2 \sqcap \dots \sqcap c_n$ stating that the class $c_1$ is equivalent to the intersection of the classes $c_2$ through $c_n$.  Similarly, a sufficient relationship block takes the form
\begin{itemize}
\item[]
  \begin{tabular}{|l|l|l|l|l|l|l|l|}\hline \textsf{sufficient} & \textsf{relationship} & (\textsf{not}) & $p$ & $c_1$ & $c_2$ \\ \hline \end{tabular} \end{itemize} where $p$ is an object property, $c_1$ and $c_2$ are classes, and \textsf{not} is an optional qualifier.  A sufficient relationship block (without the \textsf{not} qualifier) induces the description logic axiom $c_1 \equiv \exists p . c_2$ stating that the class $c_1$ is equivalent to the set of individuals that are $p$-related to at least one individual of class $c_2$. The \textsf{not} qualifier states that the absence of the property is a defining characteristic of the class. Thus, a sufficient relationship block containing a \textsf{not} qualifier induce the description logic axiom $c_1 \equiv \neg \exists p . c_2$.  The use of multiple sufficient blocks for a particular class results in a single axiom in which each constraint (i.e., class constructor) is combined via intersection ($\sqcap$). For example, the following blocks 
\begin{small}
\begin{tabbing}
~~\textsf{sufficient} \= \textsf{relationship} \= \textsf{Eutheria} \= \textsf{Mammal} \= \kill 
~~\textsf{sufficient} \> \textsf{relationship} \> \textsf{hasPart} \> \textsf{Mammal} \> \textsf{Hair} \\
  ~~\textsf{sufficient} \> \textsf{entity} \> \textsf{Eutheria} \> \textsf{Mammal} \\
  ~~\textsf{sufficient} \> \textsf{relationship} \> \textsf{not} \> \textsf{hasPart} \> \textsf{Eutheria}  \textsf{EpipubicBone} 
\end{tabbing}
induce the following description logic axioms.
\begin{tabbing}
~~$\textsf{Mammal} \equiv \exists \textsf{hasPart}.\textsf{Hair}$\\ 
~~$\textsf{Eutheria} \equiv \textsf{Mamma} \sqcap \neg \exists \textsf{hasPart} . \textsf{EpipubicBone}$
\end{tabbing}
\end{small}
These axioms state that \emph{any} individual that has Hair as a part is a Mammal and that \emph{any} Mammal that does not have an Epipubic Bone as a part is a Eutheria.

@@

\figref{fig:owlifier} shows the general architecture of the \owlifier\ application. A scientist first creates a spreadsheet containing a set of ontology definitions using \owlifier\ conventions.  The spreadsheet is then exported as a plain-text file containing the \owlifier\ table, e.g., using a CSV or tab-delimited format. The \owlifier\ table is then sent to the \owlifier\ application, which performs a number of steps that include: (i) parsing the file; (ii) generating the corresponding OWL-DL representation (e.g., in the current implementation, the OWL-API is used \citep{horridge07:_ignit_owl}, although Jena \citep{carroll04:_jena} could be used as well); (iii) validating the ontology and ensuring it is consistent via an OWL-DL reasoner (e.g., the current implementation uses Pellet \citep{sirin07:_pellet}); and (iv) outputting the corresponding OWL-DL file. The resulting OWL file can then be refined and extended via existing OWL tools (e.g., Protege or SWOOP). It is also possible for the refined ontology to be converted back to a corresponding \owlifier\ representation (shown using dashed lines in \figref{fig:owlifier}). While straightforward to perform, this ``back'' conversion is not information preserving since not all OWL-DL statements can be represented through \owlifier\ blocks.  That is, the conversion from OWL-DL to \owlifier\ will only preserve OWL-DL constructs that are represented by \owlifier\ blocks.  In particular, \owlifier\ does not support a number of OWL-DL language constructs, which include domain and range constraints, datatype properties, sub properties, property constraints other than transitive (i.e., functional, symmetric, inverse, inverse functional), value restrictions, individuals, etc., as well as various combinations of OWL-DL constructs.

The current implementation of \owlifier\ is written as an open-source Java application\footnote{See \url{https://semtools.ecoinformatics.org/owlifier}} and supports a subset of the blocks defined in \secref{sec:owlifier}. In particular, we are currently extending \owlifier\ to fully support sufficient blocks (the remaining blocks described in \secref{sec:owlifier} are implemented) as well as the potentially lossy, back conversion from OWL-DL files to corresponding \owlifier\ blocks. The current implementation of \owlifier\ is being used within a project focused on integrating vegetation trait data.  As noted above, not all the blocks described here are yet supported, however, these priority needs are clearly emerging from our interactions with the vegetation scientists.  Nevertheless, these scientists were able to rapidly prototype and revise their trait ontologies using \owlifier, with little to no instruction in formal logic or knowledge modeling.  In addition, \owlifier\ is being used to construct term hierarchies from large sets of keywords (harvested from existing metadata documents) as well as to define articulations between existing ontologies and the observation ontology defined in \citep{madin07:_ontol_for_descr_and_synth}. As we continue to use \owlifier\ within these projects, we intend to extend the application as needed to support additional blocks and services.

@@

This paper presents a new approach for developing ontologies to address barriers in ontology development and adoption by ecologists. Our approach allows scientists to use familiar spreadsheet software (e.g., frequently used by ecologists for storing and analyzing data) to create ontologies by filling in a set of templates, or blocks, that generate OWL-DL class hierarchies, properties, and constraints via the \owlifier\ application.  This approach can provide a more intuitive and accessible ontology editing environment for ecologists, especially compared to existing OWL-based tools that require considerable expertise in the underlying logic formalisms. Similar to standard OWL ontologies, \owlifier\ spreadsheets can import and extend existing OWL-DL ontologies (including those generated from \owlifier\ spreadsheets), which further supports the community development and interoperability of ontologies. Further, \owlifier\ blocks (or similarly, partially filled in blocks) can be reused during ontology creation to help ensure that future ontology development adheres to best practices and community standards (e.g., similar to ``part-of'' properties defined in the Gene Ontology \citep{bada04:_short_study_succes_of_gene_ontol}).
